---
title: "Phenotype organisation"
date: "`r Sys.Date()`"
output: html_document
---

Files needed to input into this:
 - configuration environment
 - phenotype description (name, id, units, min, max etc..)
 - genetic covariates (id, PC's, sex*)
 - phenotype file (id, age, phenotype value, covariates)
 
Minimum phenotype columns required: 'FID' 'IID' 'age' 'value' then any covariates. Value is the value of the trait, column name should be 'value' and the file name the id of the trait (e.g. 'bmi.txt' for body mass index). Phenotypic covariates (and column names) required for each trait are described in "phenotypes_list" csv file. 

*Sex can be input in either genetic covariates file or phenotype data file but is assumed to be coded as 1 - male, 2 - female.
 
Other notes:

 - If running script for multiple ancestries will need to specify a different folder for each ancestry in the enviroment file to ensure that different ancestries aren't written over each other. 
 - The code drops anything with NA rather than being specific to the variables of interest
 - Age is assumed to be in years
 - If ALL observations for a trait are removed as outliers the code will crash. Outliers are removed based on pre-transformed variables.


```{r libraries, include=FALSE, message=FALSE}
library(rms)
library(quantreg)
library(git2r)
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(here)
library(rmarkdown)
library(dotenv)
nthreads <- as.numeric(Sys.getenv("env_threads"))
```

## Set up environment

```{r environment}
source(here("resources", "phenotypes", "phenotype_functions.R"))
readRenviron(here("config.env"))
dir.create(Sys.getenv("results_dir"), showWarnings = FALSE)
dir.create(Sys.getenv("phenotype_processed_dir"), showWarnings = FALSE)
os <- list()
```

## Age parameters

```{r agebins}
agebins <- fread(here("resources", "phenotypes", "agebins.csv"), nThread = nthreads) %>% as_tibble()
```

## Phenotype parameters

```{r phenotypes}
df <- fread(here("phenotype_list.csv"), nThread = nthreads) %>% 
  as_tibble() %>%
  mutate(filepath = file.path(Sys.getenv("phenotype_input_dir"), paste0(pheno_id, ".txt"))) %>%
  filter(file.exists(filepath))
df
phenotypes <- df$pheno_id
os$phenotypes <- phenotypes
```

## Genetic covariates

```{r genetic_covariates}
gen_covs <- read_gen_covs(
  file = file.path(Sys.getenv("genotype_processed_dir"), paste0("pcs.txt")),
  npcs = as.numeric(Sys.getenv("env_n_pcs"))
)
str(gen_covs)
```

Plot PCs

```{r plot_pcs}
temp1 <- subset(gen_covs, select=-c(PC1)) %>% tidyr::pivot_longer(cols=-c("FID", "IID"))
temp2 <- subset(gen_covs, select=c(FID, IID, PC1))
temp <- inner_join(temp1, temp2)
ggplot(temp, aes(x=PC1, y=value)) +
geom_point() +
facet_wrap(~ name, scale="free")
```

## Static covariates

```{r static_covariates}
cov_list <- strsplit(Sys.getenv("static_covariates"), split=" ")[[1]]
covdat <- read_covariate_data(
  file.path(Sys.getenv("phenotype_input_dir"), "pheno_covariates.txt"),
  cov_list
)
# SENSITIVE, don't print any data
```

Age distribution

```{r age_distribution}
covdat$deob <- cut(covdat$yob/10, breaks=seq(180,203, by=1))
os$deob_summary <- table(covdat$deob)
os$deob_summary
```

Sex distribution

```{r sex_distribution}
os$sex_table <- table(covdat$sex)
os$sex_table
```

Sex x yob

```{r sex_yob}
os$sex_yob <- covdat %>% 
  group_by(sex) %>% 
  summarise(
      yob_mean = mean(yob, na.rm=T), 
      yob_sd = sd(yob, na.rm=T)
  )
os$sex_yob
saveRDS(os, file = file.path(Sys.getenv("results_dir"), "02", "summary.rds"))
```

## Processing phenotypes

- Input Phenotype data
- Remove outliers
- Summarise the data
- Save age and sex specific files for the GWAS

```{r phenotype_processing, echo=FALSE}
for(i in 1:length(phenotypes)){
  cat("\n### ", phenotypes[i], "\n\n")
  organise_phenotype(phenotypes[i], phenotypes, df, gen_covs, covdat, agebins, pl=TRUE)
}
```

## Run details

Get git version

```{r version}
g <- git2r::revparse_single(here(), "HEAD")
g
```

Session info

```{r session_info}
sessionInfo()
```
