---
title: "Phenotype organisation"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(git2r)
library(dplyr)
library(jsonlite)
library(ggplot2)
library(data.table)
library(here)
library(rmarkdown)
```

Read config file

```{r}
config <- jsonlite::read_json(here("config.json"))
```

Reading in phenotype parameters here but it will just be easier to distribute the spreadsheet with the pipeline.

```{r}
# df <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1LtsAujOKq5BK55A2BarJ_dHxOJI_4RQGwIu-ysoJxbI/edit?usp=sharing")
```


## Covariates

Read covariates file

```{r}
cs <- list()
covs <- fread(file.path(config$phenotype_input_dir, "covs.txt"))
str(covs %>% dplyr::select(-id))
```

To do: check columns

### YOB distribution

```{r}
hist(covs$yob, breaks=length(unique(round(covs$yob))))
```

```{r}
cs$yob_summary <- summary(covs$yob)
cs$yob_summary
```

Decade of birth

```{r}
covs$deob <- cut(covs$yob-1800, breaks=seq(0,300, by=10))
cs$deob_summary <- table(covs$deob)
cs$deob_summary
```

To do: check values are sensible

### Sex

```{r}
cs$sex_table <- table(covs$sex)
cs$sex_table
```

```{r}
cs$sex_yob <- covs %>% 
    group_by(sex) %>% 
    summarise(
        yob_mean = mean(yob, na.rm=T), 
        yob_sd = sd(yob, na.rm=T)
    )
cs$sex_yob
```

To do: check values are 1 / 2

### Ancestry

```{r}
cs$ancestry_table <- table(covs$ancestry)
cs$ancestry_table
```

```{r}
cs$ancestry_yob <- covs %>% 
    group_by(ancestry) %>% 
    summarise(
        yob_mean = mean(yob, na.rm=T), 
        yob_sd = sd(yob, na.rm=T)
    )
cs$ancestry_yob
```

```{r}
cs$ancestry_sex <- covs %>% 
    group_by(ancestry) %>% 
    summarise(
        sex_mean = mean(sex, na.rm=T), 
        sex_sd = sd(sex, na.rm=T)
    )
cs$ancestry_sex
```

---

## Phenotypes

```{r}
read_phenotype_data <- function(phecode, input_dir) {
    # Check if phenotype is present
    filename <- file.path(input_dir, paste0(phecode, ".txt"))
    if(!file.exists(filename)) {
        message(phecode, " not present")
        return(NULL)
    }

    # Read in phenotype
    phen <- data.table::fread(filename)

    # Check columns are there as expected
    column_names <- c("id", "age", "value")
    if(!all(column_names %in% names(phen))) {
        print(head(phen))
        stop("expect 'id', 'age' and 'value' columns to be present")
    }

    # Cut into age bins
    phen$agebin <- cut(phen$age, breaks=c(0:19, seq(20, 120, by=5)))
    return(phen)
}

plot_phen_all <- function(phen, phecode) {
    ggplot(phen, aes(x=agebin, y=value)) +
    geom_boxplot() +
    labs(title=phecode)
}

plot_phen_traj <- function(phen, phecode) {
    ggplot(phen, aes(x=agebin, y=value, group=id)) +
    geom_line(aes(group=id), alpha=0.2) +
    labs(title=phecode)
}

detect_outliers <- function() {

}

summarise_phen <- function(phen, covs) {
    dat <- inner_join(phen, covs, by="id")
    dat %>% 
        group_by(ancestry, agebin, deob) %>%
        summarise(
            n=sum(!is.na(value)), 
            m=mean(value, na.rm=T),
            sd=sd(value, na.rm=T),
            min=min(value, na.rm=T),
            max=max(value, na.rm=T),
            q5=quantile(value, 0.05, na.rm=T),
            q25=quantile(value, 0.25, na.rm=T),
            q50=quantile(value, 0.5, na.rm=T),
            q75=quantile(value, 0.75, na.rm=T),
            q95=quantile(value, 0.95, na.rm=T)
        )
}
```


### BMI

```{r}
phecode <- "bmi"
phen <- read_phenotype_data(phecode, config$phenotype_input_dir)
```

```{r eval=!is.null(phen)}
str(phen %>% dplyr::select(-id))
```

```{r eval=!is.null(phen)}
plot_phen_all(phen, phecode)
```

```{r eval=!is.null(phen)}
plot_phen_traj(phen, phecode)
```

```{r eval=!is.null(phen)}
cs[[phecode]] <- summarise_phen(phen, covs)
cs[[phecode]]
```

### LDL

```{r}
phecode <- "ldl"
phen <- read_phenotype_data(phecode, config$phenotype_input_dir)
```

```{r eval=!is.null(phen)}
str(phen %>% dplyr::select(-id))
```

```{r eval=!is.null(phen)}
plot_phen_all(phen, phecode)
```

```{r eval=!is.null(phen)}
plot_phen_traj(phen, phecode)
```

```{r eval=!is.null(phen)}
cs[[phecode]] <- summarise_phen(phen, covs)
cs[[phecode]]
```


## Save summary statistics

```{r}
dir.create(config$results_dir, recursive=T)
saveRDS(cs, file=file.path(config$results_dir, "phenotype_summaries.rds"))
```
---

## Run details

Get git version

```{r}
g <- git2r::revparse_single(here(), "HEAD")
g
```

Session info

```{r}
sessionInfo()
```